---
title: "Tumor-intrinsic mutations"
author: "mleukam"
date: "2019-07-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Clear the workspace
```{r}
rm(list = ls())
```

Load packages
```{r}
library(maftools)
library(tidyverse)
library(RColorBrewer)
library(knitr)
library(ggpubr)
library(ggsci)
```

## Read in and clean data

Read in NCI mutation data
```{r}
# read in significance rank list from MutSig2CV
mutsig_siggenes <- read_delim("data/sig_genes.txt", "\t")

# read in final analysis MAF from MutSig2CV
final_analysis_set <- read_delim(
  "data/final_analysis_set.maf", "\t")

# repair nonstandard variant classification
final_analysis_set$Variant_Classification <- str_replace(final_analysis_set$Variant_Classification, "Start_Codon_Del", "Translation_Start_Site")

final_analysis_set$Variant_Classification <- str_replace(final_analysis_set$Variant_Classification, "Stop_Codon_Ins", "Nonstop_Mutation")

final_analysis_set <- as.data.frame(final_analysis_set)

nci_mut_by_sample <- final_analysis_set %>%
  dplyr::select(sample_id = Tumor_Sample_Barcode, Hugo_Symbol) %>%
  mutate(Hugo_Symbol = ifelse(Hugo_Symbol == "MLL2", "KMT2D", Hugo_Symbol)) %>%
  as_tibble() %>%
  print()
  
```

Read in Duke mutation data
```{r}
duke_mut_raw <- read_csv("data/duke_study_driver_mutations_top150_maf.csv")

duke_mut_by_sample <- duke_mut_raw %>%
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol) %>%
  tidyr::separate(Tumor_Sample_Barcode, 
                  into = c(NA, NA, "sample_id"), 
                  sep = "_")  %>%
  mutate(sample_id = paste0("duke_", sample_id)) %>%
  arrange(sample_id) %>%
  as_tibble() %>%
  mutate(Hugo_Symbol = ifelse(Hugo_Symbol == "MLL2", "KMT2D", Hugo_Symbol)) %>%
  print()
```

#### Combine mutation data

```{r}
combined_mut <- nci_mut_by_sample %>%
  bind_rows(duke_mut_by_sample) %>%
  print()

# remove duplicate mutations in the same patient and same gene
unique_mut <- combined_mut %>%
  distinct(sample_id, Hugo_Symbol, .keep_all = TRUE) %>%
  print()

nrow(combined_mut)
nrow(unique_mut)
```

Read in cluster assignment and combine
```{r}
gsva4_es <- readRDS("output/combined_es_gsva4_4cluster.rds")

clust_assgn <- pData(gsva4_es) %>% 
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  dplyr::select(sample_id, clust) %>%
  as_tibble() %>%
  print()

# get the total n
clust_assgn %>% distinct(sample_id) %>% nrow()

mut_clust <- unique_mut %>%
  left_join(clust_assgn) %>%
  print()
```

#### Mutation counts by cluster

Get mutation counts per gene for each cluster and combine back to a single table
```{r}
# get number of cases in the cluster
mut_clust %>% dplyr::filter(clust == "1") %>% distinct(sample_id) %>% nrow()
mut_clust %>% dplyr::filter(clust == "2") %>% distinct(sample_id) %>% nrow()
mut_clust %>% dplyr::filter(clust == "3") %>% distinct(sample_id) %>% nrow()
mut_clust %>% dplyr::filter(clust == "4") %>% distinct(sample_id) %>% nrow()

# get counts by gene for clusters and compare each cluster for significance against pooled counts from each other cluster

# first get mutation counts for each gene per cluster and create column for counts of unmutated genes
cluster_1 <- mut_clust %>%
  dplyr::filter(clust == "1") %>%
  dplyr::select(-clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c1_mut = n) %>%
  mutate(c1_nonmut = 352 - c1_mut) %>%
  print()

cluster_2 <- mut_clust %>%
  dplyr::filter(clust == "2") %>%
  dplyr::select(-clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c2_mut = n) %>%
  mutate(c2_nonmut = 277 - c2_mut) %>%
  print() 

cluster_3 <- mut_clust %>%
  dplyr::filter(clust == "3") %>%
  dplyr::select(-clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c3_mut = n) %>%
  mutate(c3_nonmut = 222 - c3_mut) %>%
  print() 

cluster_4 <- mut_clust %>%
  dplyr::filter(clust == "4") %>%
  dplyr::select(-clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c4_mut = n) %>%
  mutate(c4_nonmut = 317 - c4_mut) %>%
  print() 

mut_counts_by_cluster <- cluster_1 %>%
  full_join(cluster_2) %>%
  full_join(cluster_3) %>%
  full_join(cluster_4) %>%
  replace_na(list(c1_mut = 0, c2_mut = 0, c3_mut = 0, c4_mut = 0)) %>%
  replace_na(list(c1_nonmut = 0, c2_nonmut = 0, c3_nonmut = 0, c4_nonmut = 0)) %>%
  arrange(desc(c4_mut)) %>%
  print()

mut_counts_matrix <- as.data.frame(mut_counts_by_cluster) %>%
  column_to_rownames(var = "Hugo_Symbol") %>%
  as.matrix()

mut_counts_matrix[1:5, ]
mut_counts <- mut_counts_matrix[, c(1,3,5,7)]
  
# filter for genes without any meaningful number of total mutations
# filter out genes that have a mutant allele fraction < 0.05
cutoff <- 1189 * 0.05

mut_counts_filtered <- mut_counts_by_cluster %>%
  mutate(mut_totals = c1_mut + c2_mut + c3_mut + c4_mut) %>%
  filter(mut_totals > cutoff) %>%
  print()
```

## Test pairwise significance

Significance testing per for each cluster vs pool of other clusters

https://stackoverflow.com/questions/41507230/fisher-test-on-multiple-data-entries-and-extract-results-in-separate-table/41512086#41512086?newreg=fbae75b1bdbc4796b925c53ca1e2f132

https://blog.rstudio.com/2016/02/02/tidyr-0-4-0/

https://jennybc.github.io/purrr-tutorial/index.html

```{r}
clust1_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c2_mut + c3_mut + c4_mut) %>%
  mutate(pool_nonmut = c2_nonmut + c3_nonmut + c4_nonmut) %>%
  dplyr::select(Hugo_Symbol, c1_mut, c1_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust1_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_1_summary <- sig %>%
  left_join(clust1_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  print()
```

```{r}
clust2_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c1_mut + c3_mut + c4_mut) %>%
  mutate(pool_nonmut = c1_nonmut + c3_nonmut + c4_nonmut) %>%
  dplyr::select(Hugo_Symbol, c2_mut, c2_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust2_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_2_summary <- sig %>%
  left_join(clust2_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  print()
```

```{r}
clust3_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c1_mut + c2_mut + c4_mut) %>%
  mutate(pool_nonmut = c1_nonmut + c2_nonmut + c4_nonmut) %>%
  dplyr::select(Hugo_Symbol, c3_mut, c3_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust3_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_3_summary <- sig %>%
  left_join(clust3_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  print()
```

```{r}
clust4_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c1_mut + c2_mut + c3_mut) %>%
  mutate(pool_nonmut = c1_nonmut + c2_nonmut + c3_nonmut) %>%
  dplyr::select(Hugo_Symbol, c4_mut, c4_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust4_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_4_summary <- sig %>%
  left_join(clust4_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  print()
```


## g:Profiler input

Get list of genes mutated over allele fraction of 0.05 for each individual cluster for input into g:profiler ordered enrichment test

Rank by odds of mutation

```{r}
cluster_1_maf05 <- mut_clust %>%
  dplyr::filter(clust == "1") %>%
  dplyr::select(-clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c1_mut = n) %>%
  mutate(MAF = c1_mut/352) %>%
  dplyr::filter(MAF > 0.01) %>%
  print()

mut_clust_c1_maf05 <- mut_clust %>%
  left_join(cluster_1_maf05) %>%
  print()
  
clust1_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c2_mut + c3_mut + c4_mut) %>%
  mutate(pool_nonmut = c2_nonmut + c3_nonmut + c4_nonmut) %>%
  dplyr::select(Hugo_Symbol, c1_mut, c1_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust1_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_1_summary <- sig %>%
  left_join(clust1_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  print()
```


