---
title: "new_deconvolution_results"
author: "mleukam"
date: "2020-02-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Clear workspace
```{r}
rm(list = ls())
```

Prevent knitting from executing chunks
```{r}
knitr::opts_chunk$set(eval = FALSE)
```

Load packages
```{r}
# plotting
library(ggsci)
library(ggpubr)
library(viridis)
library(ComplexHeatmap)
library(circlize)
library(scales)
library(ggConvexHull)

# correlation
library(corrplot)
source("http://www.sthda.com/upload/rquery_cormat.r")

# clustering
library(FactoMineR)
library(factoextra)
library(cluster)

# survival analysis
library(survival)
library(survminer)

# utility
library(rlang)
library(broom)
library(Biobase)
library(stats)

# base tidyverse
library(tidyverse)
```


## Read in data

```{r}
nci_decon <- read_csv("data/cibersort/NCI/abs_cell_number_estimate_022320/CIBERSORTx_Job2_Results.csv")

duke_decon <- read_csv("data/cibersort/Duke/abs_cell_number_estimate_022320/CIBERSORTx_Job3_Results.csv")

combined_decon <- nci_decon %>%
  bind_rows(duke_decon) %>%
  print()
```


## Boxplots by GSVA cluster

Read in cluster assignment and combine
```{r}
gsva4_es <- readRDS("output/combined_es_gsva4_4cluster.rds")

gsva_clust_assgn <- pData(gsva4_es) %>% 
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  dplyr::select(sample_id, clust) %>%
  as_tibble() %>%
  print()

decon_res <- combined_decon %>%
  rename(sample_id = Mixture) %>%
  left_join(gsva_clust_assgn) %>%
  dplyr::select(sample_id, clust, everything()) %>%
  print()

cd8_plot <- ggplot(decon_res, 
                   aes(x = clust, y = cd8_eff, fill = clust)) +
  geom_boxplot() +
    theme_classic2() + 
        scale_fill_d3(name = "Cluster", labels = c("Cold",   "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("CD8 Deconvolution Score") +
  stat_compare_means() +
  ggtitle("CD8")
cd8_plot

th1_plot <- ggplot(decon_res, 
                   aes(x = clust, y = th1_rest, fill = clust)) +
  geom_boxplot() +
    theme_classic2() + 
        scale_fill_d3(name = "Cluster", labels = c("Cold",   "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Th1 Deconvolution Score") +
  stat_compare_means() +
  ggtitle("Th1")
th1_plot

th2_plot <- ggplot(decon_res, 
                   aes(x = clust, y = th2_rest, fill = clust)) +
  geom_boxplot() +
    theme_classic2() + 
        scale_fill_d3(name = "Cluster", labels = c("Cold",   "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Th2 Deconvolution Score") +
  stat_compare_means() +
  ggtitle("Th2")
th2_plot

mac_1_plot <- ggplot(decon_res, 
                   aes(x = clust, y = mac_1, fill = clust)) +
  geom_boxplot() +
    theme_classic2() + 
      scale_fill_d3(name = "Cluster", labels = c("Cold",   "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Macrophage Type 1 Deconvolution Score") +
  stat_compare_means() +
  ggtitle("Macrophage Type 1")
mac_1_plot

mac_2_plot <- ggplot(decon_res, 
                   aes(x = clust, y = mac_2, fill = clust)) +
  geom_boxplot() +
    theme_classic2() + 
      scale_fill_d3(name = "Cluster", labels = c("Cold",   "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Macrophage Type 2 Deconvolution Score") +
   stat_compare_means() +
    ggtitle("Macrophage Type 2")
mac_2_plot

mac_3_plot <- ggplot(decon_res, 
                   aes(x = clust, y = mac_3, fill = clust)) +
  geom_boxplot() +
    theme_classic2() + 
    scale_fill_d3(name = "Cluster", labels = c("Cold",   "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Macrophage Type 3 Deconvolution Score") +
   stat_compare_means() +
    ggtitle("Macrophage Type 3")
mac_3_plot

mac_4_plot <- ggplot(decon_res, 
                   aes(x = clust, y = mac_4, fill = clust)) +
  geom_boxplot() +
    theme_classic2() + 
  scale_fill_d3(name = "Cluster", labels = c("Cold",   "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Macrophage Type 4 Deconvolution Score") +
   stat_compare_means() +
    ggtitle("Macrophage Type 4")
mac_4_plot

mac_5_plot <- ggplot(decon_res, 
                   aes(x = clust, y = mac_5, fill = clust)) +
  geom_boxplot() +
  theme_classic2() + 
  scale_fill_d3(name = "Cluster", labels = c("Cold", "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Macrophage Type 5 Deconvolution Score") +
   stat_compare_means() +
    ggtitle("Macrophage Type 5")
mac_5_plot

mac_6_plot <- ggplot(decon_res, 
                   aes(x = clust, y = mac_6, fill = clust)) +
  geom_boxplot() +
  theme_classic2() + 
  scale_fill_d3(name = "Cluster", labels = c("Cold", "Intermediate-T", "Intermediate-M", "Inflamed")) +
  xlab("") +
  ylab("Macrophage Type 6 Deconvolution Score") +
   stat_compare_means() +
    ggtitle("Macrophage Type 6")
mac_6_plot
```


## Survival analysis

Read in clinical data
```{r}
nci_pheno_data <- read_csv("output/nci_extended_pheno_data.csv") %>% print()

nci_surv_frame <- nci_pheno_data %>%
  dplyr::select(sample_id,
                os_status = `Status at Follow_up_ 0 Alive_ 1 Dead`,
                os_years = `Follow_up Time _yrs`,
                pfs_status = `Progression_Free Survival _PFS_ Status_ 0 No Progressoin_ 1 Progression`,
                pfs_years = `Progression_Free Survival _PFS_ Time _yrs`) %>%
  print()

nci_extremes <- nci_decon %>%
  mutate(cd8_groups = ntile(cd8_eff, 5)) %>%
  mutate(th1_groups = ntile(th1_rest, 5)) %>%
  mutate(th2_groups = ntile(th2_rest, 5)) %>%
  mutate(treg_groups = ntile(treg, 5)) %>%
  mutate(nk_groups = ntile(nk_rest, 5)) %>%
  mutate(mac1_groups = ntile(mac_1, 5)) %>%
  mutate(mac2_groups = ntile(mac_2, 5)) %>%
  mutate(mac3_groups = ntile(mac_3, 5)) %>%
  mutate(mac4_groups = ntile(mac_4, 5)) %>%
  mutate(mac5_groups = ntile(mac_5, 5)) %>%
  mutate(mac6_groups = ntile(mac_6, 5)) %>%
  mutate(mac7_groups = ntile(mac_7, 5)) %>%
  dplyr::select(sample_id = Mixture, cd8_groups, th1_groups, th2_groups, treg_groups, nk_groups, mac1_groups, mac2_groups, mac3_groups, mac4_groups, mac5_groups, mac6_groups, mac7_groups) %>%
  print()

nci_surv_frame_groups <- nci_surv_frame %>%
  left_join(nci_extremes) %>%
  print()

nci_cd8_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(cd8_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, cd8_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_cd8_os_object <- Surv(time = nci_cd8_extremes$os_years,
                   event = nci_cd8_extremes$os_status)
fit_cd8 <- survfit(nci_cd8_os_object ~ cd8_groups,
                   data = nci_cd8_extremes)
ggsurvplot(fit_cd8, data = nci_cd8_extremes, pval = TRUE)

nci_th1_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(th1_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, th1_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_th1_os_object <- Surv(time = nci_th1_extremes$os_years,
                   event = nci_th1_extremes$os_status)
fit_th1 <- survfit(nci_th1_os_object ~ th1_groups,
                   data = nci_th1_extremes)
ggsurvplot(fit_th1, data = nci_th1_extremes, pval = TRUE)

nci_th2_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(th2_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, th2_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_th2_os_object <- Surv(time = nci_th2_extremes$os_years,
                   event = nci_th2_extremes$os_status)
fit_th2 <- survfit(nci_th2_os_object ~ th2_groups,
                   data = nci_th2_extremes)
ggsurvplot(fit_th2, data = nci_th2_extremes, pval = TRUE)

nci_treg_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(treg_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, treg_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_treg_os_object <- Surv(time = nci_treg_extremes$os_years,
                   event = nci_treg_extremes$os_status)
fit_treg <- survfit(nci_treg_os_object ~ treg_groups,
                   data = nci_treg_extremes)
ggsurvplot(fit_treg, data = nci_treg_extremes, pval = TRUE)

nci_mac1_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(mac1_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, mac1_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_mac1_os_object <- Surv(time = nci_mac1_extremes$os_years,
                   event = nci_mac1_extremes$os_status)
fit_mac1 <- survfit(nci_mac1_os_object ~ mac1_groups,
                   data = nci_mac1_extremes)
ggsurvplot(fit_mac1, data = nci_mac1_extremes, pval = TRUE)

nci_mac2_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(mac2_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, mac2_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_mac2_os_object <- Surv(time = nci_mac2_extremes$os_years,
                   event = nci_mac2_extremes$os_status)
fit_mac2 <- survfit(nci_mac2_os_object ~ mac2_groups,
                   data = nci_mac2_extremes)
ggsurvplot(fit_mac2, data = nci_mac2_extremes, pval = TRUE)

nci_mac3_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(mac3_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, mac3_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_mac3_os_object <- Surv(time = nci_mac3_extremes$os_years,
                   event = nci_mac3_extremes$os_status)
fit_mac3 <- survfit(nci_mac3_os_object ~ mac3_groups,
                   data = nci_mac3_extremes)
ggsurvplot(fit_mac3, data = nci_mac3_extremes, pval = TRUE)

nci_mac4_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(mac4_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, mac4_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_mac4_os_object <- Surv(time = nci_mac4_extremes$os_years,
                   event = nci_mac4_extremes$os_status)
fit_mac4 <- survfit(nci_mac4_os_object ~ mac4_groups,
                   data = nci_mac4_extremes)
ggsurvplot(fit_mac4, data = nci_mac4_extremes, pval = TRUE)

nci_mac5_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(mac5_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, mac5_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_mac5_os_object <- Surv(time = nci_mac5_extremes$os_years,
                   event = nci_mac5_extremes$os_status)
fit_mac5 <- survfit(nci_mac5_os_object ~ mac5_groups,
                   data = nci_mac5_extremes)
ggsurvplot(fit_mac5, data = nci_mac5_extremes, pval = TRUE)

nci_mac6_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(mac6_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, mac6_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_mac6_os_object <- Surv(time = nci_mac6_extremes$os_years,
                   event = nci_mac6_extremes$os_status)
fit_mac6 <- survfit(nci_mac6_os_object ~ mac6_groups,
                   data = nci_mac6_extremes)
ggsurvplot(fit_mac6, data = nci_mac6_extremes, pval = TRUE)

nci_mac7_extremes <- nci_surv_frame_groups %>%
  dplyr::filter(mac7_groups %in% c("1", "5")) %>%
  dplyr::select(sample_id, mac7_groups, os_years, os_status, pfs_years, pfs_status) %>%
  print()
nci_mac7_os_object <- Surv(time = nci_mac7_extremes$os_years,
                   event = nci_mac7_extremes$os_status)
fit_mac7 <- survfit(nci_mac7_os_object ~ mac7_groups,
                   data = nci_mac7_extremes)
ggsurvplot(fit_mac7, data = nci_mac7_extremes, pval = TRUE)
```

## Mutations
```{r}
# rename nci decon columns
nci_decons <- nci_decon %>% 
  rename(sample_id = Mixture) %>%
  print()

# collapse columns to larger groups


# get column names
decon_names <- colnames(nci_decons)[2:27]
decon_names

# DLBCL mutation data
muts_dlbcl <- read_delim("data/final_analysis_set.maf", 
    "\t", escape_double = FALSE, trim_ws = TRUE) %>%
  dplyr::select(Tumor_Sample_Barcode, everything()) %>%
  dplyr::select(sample_id = Tumor_Sample_Barcode, 
                Hugo_Symbol) %>%
  distinct(sample_id, Hugo_Symbol) %>%
  dplyr::mutate(n = rep(1, length.out = nrow(.))) %>%
  spread(key = Hugo_Symbol, value = n, fill = 0) %>%
  as.data.frame() %>%
  column_to_rownames(var = "sample_id") %>%
  as.matrix()
muts_dlbcl[1:5, 1:5]
str(muts_dlbcl)

# keep only those columns with >5% mutations
muts_5 <- muts_dlbcl[, colSums(muts_dlbcl) > 24]
str(muts_5)
# reduces to 96 candidate genes

muts_5_df <- muts_5 %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble() %>%
  print()

mut_names <- colnames(muts_5_df)[-1]
mut_names

# get all permutations of selected mutations and deconvolution
perms <- expand.grid(mut_names, decon_names) %>%
  rename(mut_names = Var1, decon_names = Var2) %>%
  mutate(mut_names = as.character(mut_names)) %>%
  mutate(decon_names = as.character(decon_names))
str(perms)

# write function to compare deconvolution scores by mutation
decon_vs_mut <- function(mut_name, decon_name){
  mut <- muts_5_df %>%
    filter(!! rlang::sym(mut_name) == 1) %>%
    select(sample_id) %>%
    left_join(nci_decons) %>%
    pull(!! rlang::sym(decon_name))
  nonmut <- muts_5_df %>%
    filter(!! rlang::sym(mut_name) == 0) %>%
    select(sample_id) %>%
    left_join(nci_decons) %>%
    pull(!! rlang::sym(decon_name))
  tidy(t.test(mut, nonmut)) %>%
    bind_cols("gene" = mut_name) %>%
    bind_cols("cell_type" = decon_name) %>%
    bind_cols("mut_number" = length(mut)) %>%
    bind_cols("nonmut_number" = length(nonmut)) %>%
    mutate(mut_rate = length(mut) / (length(mut) + length(nonmut))) %>%
    mutate(mut_mean = sum(mut, na.rm = TRUE) / length(mut)) %>%
    mutate(nonmut_mean = sum(nonmut, na.rm = TRUE) / length(nonmut)) %>%
    mutate(change = (mut_mean - nonmut_mean)) %>%
    select(gene, cell_type, p.value, change, mut_mean, nonmut_mean, mut_rate, mut_number, nonmut_number, everything())
}

# test function
decon_vs_mut("EZH2", "cd8_eff")

mutation_correlations <- map2_dfr(perms$mut_names, 
                                  perms$decon_names, 
                                  decon_vs_mut)

mut_decon_corr <- mutation_correlations %>%
  arrange(p.value) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(gene, cell_type, p.value, p.adj, everything()) %>%
  print()
  
mut_decon_corr %>% 
  write_csv("output/NCI_mutations_vs_deconvolution.csv")
```

## Validate results
```{r}


```


## Clustering by deconvolution - Abandoned

Roughly following: https://www.nature.com/articles/s41467-018-05570-1#Sec12

Correlation
```{r}
nci_score_mat <- nci_decon %>% 
  dplyr::select(-Mixture, -`P-value`, -Correlation, -RMSE, -`Absolute score (sig.score)`) %>%
  as.matrix()
rownames(nci_score_mat) <- nci_results$Mixture
nci_score_mat[1:5, 1:5]
dim(nci_score_mat)
summary(nci_score_mat)
# estimate correlation with each of the other cell types
rquery.cormat(nci_score_mat)

cormat <- rquery.cormat(nci_score_mat, graphType = "heatmap")
cormat
```

Clustering
```{r}
set.seed(818)
PCA(nci_score_mat, scale.unit = TRUE, ncp = 5, graph = TRUE)
res.pca <- PCA(nci_score_mat, graph = FALSE)
eig.val <- get_eigenvalue(res.pca)
eig.val
fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))
fviz_pca_var(res.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )

var <- get_pca_var(res.pca)
var

corrplot(var$cos2, is.corr = FALSE)
fviz_contrib(res.pca, choice = "var", axes = 1)
fviz_contrib(res.pca, choice = "var", axes = 2)

fviz_pca_ind(res.pca)

ind.p <- fviz_pca_ind(res.pca, geom = "point")
ggpubr::ggpar(ind.p,
              title = "Principal Component Analysis",
              caption = "Source: factoextra",
              xlab = "PC1", ylab = "PC2",
              ggtheme = theme_gray(), palette = "jco"
              )

# what is the optimal number of clusters?
fviz_nbclust(nci_score_mat, FUNcluster = cluster::pam, method = "gap_stat") 
fviz_nbclust(nci_score_mat, FUNcluster = cluster::pam, method = "silhouette") 
fviz_nbclust(nci_score_mat, FUNcluster = cluster::pam, method = "wss") 
```

Heatmap
```{r}
scale_mat <- nci_score_mat %>% 
  apply(1, rescale)
scale_mat[1:5, 1:5]
f1 = colorRamp2(seq(min(t(scale_mat)), max(t(scale_mat)), length = 3), c("blue", "#EEEEEE", "red"))
hm <- Heatmap(t(scale_mat),
              col = f1,
              show_row_names = FALSE,
              heatmap_legend_param = list(title = "Relative\nCell\nFraction")
              )
hm
```

Partioning clustering
http://www.sthda.com/english/wiki/the-ultimate-guide-to-partitioning-clustering

```{r}
pam.res <- pam(nci_score_mat, 3)
print(pam.res)
fviz_cluster(pam.res, 
             stand = TRUE, 
             axes = c(1:2), 
             geom = "point", 
             show.clust.cent = FALSE) +
  theme_classic2() +
  scale_color_d3() +
  scale_fill_d3()
fviz_cluster(pam.res, stand = TRUE, axes = c(1,3), geom = "point", show.clust.cent = FALSE)
pam.res$clustering
```

Build phenodata and pull in GSVA results
```{r}
gs_matrix <- readRDS("output/dlbcl_total_immune_gset_v4_results.rds")
gs_matrix[, 1:5]
str(gs_matrix)

combined_es <- readRDS("output/combined_expressionset.rds")
pheno_data <- pData(combined_es)

nci_pdata <- pData(combined_es) %>%
  rownames_to_column(var = "sample_id") %>%
  print()
cluster_pdata <- pam.res$clustering %>%
  enframe() %>%
  dplyr::rename(sample_id = name, pam_clust = value) %>%
  left_join(nci_pdata, by = "sample_id") %>%
  print()
```

Make comparisons between my clusters and ABC/GCB
```{r}
table(cluster_pdata$pam_clust, cluster_pdata$abc_gcb_rna)
table(cluster_pdata$`Genetic Subtype`, cluster_pdata$pam_clust)

cpdata_tbl <- cluster_pdata %>%
  as_tibble() %>%
  mutate(pam_clust = as.factor(pam_clust)) %>%
  dplyr::rename(pfs_status = `Progression_Free Survival _PFS_ Status_ 0 No Progressoin_ 1 Progression`) %>%
  dplyr::rename(pfs_years = `Progression_Free Survival _PFS_ Time _yrs`) %>%
  print()

library(table1)
table1(~ factor(gender) +
       age +
       factor(ipi_initial) +
       factor(ann_arbor_clinical_stage) +
       factor(abc_gcb_rna) +
       factor(pdl1_status) +
       factor(`Genetic Subtype`) | pam_clust, data = cpdata_tbl)

library(survival)
library(survminer)

# OS
# set up survival object
surv_os <- survfit(Surv(os_years, os_status) ~ pam_clust, data = cpdata_tbl)
# plot
plot1 <- ggsurvplot(surv_os, data = as.data.frame(cpdata_tbl), pval = TRUE, palette = "nejm") +
  ylab("OS probability") +
  xlab("Time (years)")

plot2 <- ggsurvplot_facet(surv_os, 
                          data = as.data.frame(cpdata_tbl), 
                          facet.by = c("abc_gcb_rna"), 
                          pval = TRUE, 
                          palette = "nejm") +
  ylab("OS probability") +
  xlab("Time (years)")
plot2

# PFS
surv_pfs <- survfit(Surv(pfs_years, pfs_status) ~ pam_clust, data = cpdata_tbl)
ggsurvplot(surv_pfs, data = cpdata_tbl, pval = TRUE, palette = "nejm") +
  ylab("PFS probability") +
  xlab("Time (years)")

plot3 <- ggsurvplot_facet(surv_pfs, 
                          data = as.data.frame(cpdata_tbl), 
                          facet.by = c("abc_gcb_rna"), 
                          pval = TRUE, 
                          palette = "nejm") +
  ylab("PFS probability") +
  xlab("Time (years)")
plot3
```

Plot scores on cluster map
```{r}
tpm_cyt_es <- readRDS("output/tpm_cyt_expressionset.rds")
tpm_pheno_data <- pData(tpm_cyt_es) %>% 
  rownames_to_column(var = "sample_id") %>% 
  as_tibble() %>% 
  print()
tpm_matrix <- exprs(tpm_cyt_es)

nci_results_ref <- nci_decon %>% 
  dplyr::rename(sample_id = Mixture)

pdata <- cluster_pdata %>% 
  left_join(tpm_pheno_data, by = "sample_id") %>%
  left_join(nci_results_ref) %>%
  mutate(pam_clust = as.factor(pam_clust)) %>%
  print()


# CYT score
fviz_pca_ind(res.pca, 
             col.ind = log(pdata$CYT),
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "CYT score\n(log scale)") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "CYT score by cluster") +
  scale_color_viridis()

# CYT score
fviz_pca_ind(res.pca, 
             col.ind = log(pdata$CYT),
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "CYT score\n(log scale)") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "CYT score by cluster") +
  scale_color_viridis()

# CD8 activation
fviz_pca_ind(res.pca, 
             col.ind = pdata$mskcc_cd8activation_gs,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "MSKCC CD8 \nactivation score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "MSKCC CD8 activation score by cluster") +
  scale_color_viridis()

# t cell anergy score
fviz_pca_ind(res.pca, 
             col.ind = pdata$mskcc_anergy_gs,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "MSKCC \nanergy score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "MSKCC T cell anergy score by cluster") +
  scale_color_viridis()

# IFN-gamma activity
fviz_pca_ind(res.pca, 
             col.ind = pdata$thor_ifng_score_21050467,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "IFNg score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "IFNg gene set score by cluster") +
  scale_color_viridis()

# CSF1 response
thor_csf1_response

fviz_pca_ind(res.pca, 
             col.ind = pdata$thor_csf1_response,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "CSF1 \nresponse score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "CSF1 response score by cluster") +
  scale_color_viridis()

# CSF1 response
fviz_pca_ind(res.pca, 
             col.ind = pdata$imsig_macrophages_gs,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "Macrophage gene\nset score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "Macrophage gene set score by cluster") +
  scale_color_viridis()

# CSF1 response
fviz_pca_ind(res.pca, 
             col.ind = pdata$imsig_macrophages_gs,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "Macrophage gene\nset score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "Macrophage gene set score by cluster") +
  scale_color_viridis()

# CSF1 response
fviz_pca_ind(res.pca, 
             col.ind = pdata$mskcc_hypoxia_regulated_gs,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "TGFb gene\nset score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "TGFb gene set score by cluster") +
  scale_color_viridis()
```

Deconvolution results on cluster map
```{r}
fviz_pca_ind(res.pca, 
             col.ind = pdata$cd8_eff,
             label = "none",
             addEllipses = FALSE,
             ggtheme = theme_classic2()) +
  labs(color = "CD8_eff score") +
  geom_convexhull(alpha = 0.0, 
                  aes(color = pdata$pam_clust,
                      fill = pdata$pam_clust),
                  show.legend = FALSE,
                  color = "black",
                  linetype = 3) +
  labs(title = "CD8_eff deconvolution by cluster") +
  scale_color_viridis()
```

Boxplots
```{r}
pdata_decon <- pdata %>% dplyr::select(1, 2, 273:298) %>%
  gather(key = cell_type, value = score, -sample_id, -pam_clust) %>%
  mutate(cell_type = as.factor(cell_type)) %>%
  print()

pdata_decon_selected <- pdata_decon %>%
  dplyr::filter(cell_type %in% c("cd8_eff", "cd8_mem", "mac_2", "mac_3", "mac_5", "mac_7", "th1_act", "th1_rest", "th2_act", "th2_rest")) %>%
  mutate(cell_type = fct_drop(cell_type)) %>%
  print()

box1 <- ggplot(pdata_decon, aes(x = cell_type, y = score, fill = cell_type)) +
  geom_boxplot() +
  facet_grid(facets = "pam_clust") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 90))
box1

box2 <- ggplot(pdata_decon_selected, aes(x = cell_type, y = score, fill = cell_type)) +
  geom_boxplot() +
  facet_grid(facets = "pam_clust") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 90))
box2

pdata_decon_selected %>% dplyr::filter(cell_type == "cd8_eff") %>% ggplot(aes(x = cell_type, y = score, fill = pam_clust)) +
  geom_boxplot() +
  facet_wrap(facets = "pam_clust") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 90))

pdata_decon_selected %>% dplyr::filter(cell_type == "mac_3") %>% ggplot(aes(x = cell_type, y = score, fill = pam_clust)) +
  geom_boxplot() +
  facet_wrap(facets = "pam_clust") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 90))

pdata_decon_selected %>% dplyr::filter(cell_type == "th2_rest") %>% ggplot(aes(x = cell_type, y = score, fill = pam_clust)) +
  geom_boxplot() +
  facet_wrap(facets = "pam_clust") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 90))
  
View(pdata_decon_selected %>% group_by(pam_clust, cell_type) %>% summarize(mean = mean(score)))
```


comparing to UCH results
```{r}
uchi_results <- read_csv("data/cibersort/UChi/absolute_cell_number_estimate_22320/CIBERSORTx_Job4_Results.csv") 

cell_seg_df <- readRDS("output/uchi_deconvolution_batch_1.rds") %>%
  separate(sample, into = c("number", "initials", sep = "-")) %>%
  mutate(sample_id = paste0("JG", number)) %>%
  dplyr::select(sample_id, everything()) %>%
  dplyr::select(-number, -initials, -`-`) %>%
  dplyr::mutate(phenotype = ifelse(phenotype == "CTL_CD8+", "cd8_eff", phenotype)) %>%
  print()

cell_density <- cell_seg_df %>%
  dplyr::select(sample_id, phenotype, cell_density_per_mm) %>%
  print()

uchi_results_df_cd8 <- uchi_results %>%
  gather(key = phenotype, value = score, -Mixture) %>%
  dplyr::rename(sample_id = Mixture) %>%
  left_join(cell_density) %>%
  dplyr::filter(phenotype == "cd8_eff") %>%
  drop_na(cell_density_per_mm) %>%
  print()

ggplot(data = uchi_results_df_cd8, aes(x = score, y = cell_density_per_mm)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic2()
```

Get predicted clusters
```{r}
uchi_score_mat <- uchi_results %>% 
  dplyr::select(-Mixture, -`P-value`, -Correlation, -RMSE, -`Absolute score (sig.score)`) %>%
  as.matrix()
rownames(uchi_score_mat) <- uchi_results$Mixture
uchi_score_mat[1:5, 1:5]
dim(uchi_score_mat)
summary(uchi_score_mat)

# object: an object of class PCA
# newdata: A data frame or a matrix in which to look for variables with which to predict.
# newdata must contain columns with the same names as the original data.
# usage: predict(object, newdata, ...)
set.seed(818)
pred_uchi <- predict(res.pca, uchi_score_mat)

# extract values for plotting
uchi_gs_df <- uchi_score_mat %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  as_tibble()
uchi_gs_df$pc1 <- pred_uchi$coord[, 1]
uchi_gs_df$pc2 <- pred_uchi$coord[, 2] 


fviz_cluster(pam.res, 
             stand = TRUE, 
             axes = c(1:2), 
             geom = "point", 
             show.clust.cent = FALSE) +
  theme_classic2() +
  scale_color_d3() +
  scale_fill_d3() +
  geom_point(data = uchi_gs_df, 
             aes(x = pc1, y = pc2), 
             color = "red", 
             shape = 19, 
             size = 2.5) + 
  geom_label(data = uchi_gs_df, aes(x = pc1, y = pc2, label = uchi_gs_df$sample_id))
```

Mutations
Read in NCI mutation data
```{r}
# read in significance rank list from MutSig2CV
mutsig_siggenes <- read_delim("data/sig_genes.txt", "\t")

# read in final analysis MAF from MutSig2CV
final_analysis_set <- read_delim(
  "data/final_analysis_set.maf", "\t")

# repair nonstandard variant classification
final_analysis_set$Variant_Classification <- str_replace(final_analysis_set$Variant_Classification, "Start_Codon_Del", "Translation_Start_Site")

final_analysis_set$Variant_Classification <- str_replace(final_analysis_set$Variant_Classification, "Stop_Codon_Ins", "Nonstop_Mutation")

final_analysis_set <- as.data.frame(final_analysis_set)

nci_mut_by_sample <- final_analysis_set %>%
  dplyr::select(sample_id = Tumor_Sample_Barcode, Hugo_Symbol) %>%
  mutate(Hugo_Symbol = ifelse(Hugo_Symbol == "MLL2", "KMT2D", Hugo_Symbol)) %>%
  as_tibble() %>%
  print()
  
# remove duplicate mutations in the same patient and same gene
unique_mut <- nci_mut_by_sample %>%
  distinct(sample_id, Hugo_Symbol, .keep_all = TRUE) %>%
  print()

clust_assgn <- pdata %>% 
  dplyr::select(sample_id, pam_clust) %>%
  as_tibble() %>%
  print()

# get the total n
clust_assgn %>% distinct(sample_id) %>% nrow()

mut_clust <- unique_mut %>%
  left_join(clust_assgn) %>%
  print()

# Get mutation counts per gene for each cluster and combine back to a single table

# get number of cases in the cluster
mut_clust %>% dplyr::filter(pam_clust == "1") %>% distinct(sample_id) %>% nrow()
mut_clust %>% dplyr::filter(pam_clust == "2") %>% distinct(sample_id) %>% nrow()
mut_clust %>% dplyr::filter(pam_clust == "3") %>% distinct(sample_id) %>% nrow()

# get counts by gene for clusters and compare each cluster for significance against pooled counts from each other cluster

# first get mutation counts for each gene per cluster and create column for counts of unmutated genes
cluster_1 <- mut_clust %>%
  dplyr::filter(pam_clust == "1") %>%
  dplyr::select(-pam_clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c1_mut = n) %>%
  mutate(c1_nonmut = 279 - c1_mut) %>%
  print()

cluster_2 <- mut_clust %>%
  dplyr::filter(pam_clust == "2") %>%
  dplyr::select(-pam_clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c2_mut = n) %>%
  mutate(c2_nonmut = 119 - c2_mut) %>%
  print() 

cluster_3 <- mut_clust %>%
  dplyr::filter(pam_clust == "3") %>%
  dplyr::select(-pam_clust) %>%
  mutate(sample_id = as.factor(sample_id),
         Hugo_Symbol = as.factor(Hugo_Symbol)) %>%
  count(Hugo_Symbol) %>%
  dplyr::rename(c3_mut = n) %>%
  mutate(c3_nonmut = 80 - c3_mut) %>%
  print() 

mut_counts_by_cluster <- cluster_1 %>%
  full_join(cluster_2) %>%
  full_join(cluster_3) %>%
  replace_na(list(c1_mut = 0, c2_mut = 0, c3_mut = 0)) %>%
  replace_na(list(c1_nonmut = 0, c2_nonmut = 0, c3_nonmut = 0)) %>%
  arrange(desc(c3_mut)) %>%
  print()

mut_counts_matrix <- as.data.frame(mut_counts_by_cluster) %>%
  column_to_rownames(var = "Hugo_Symbol") %>%
  as.matrix()

cutoff <- 1189 * 0.05

mut_counts_filtered <- mut_counts_by_cluster %>%
  mutate(mut_totals = c1_mut + c2_mut + c3_mut) %>%
  filter(mut_totals > cutoff) %>%
  print()

clust1_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c2_mut + c3_mut) %>%
  mutate(pool_nonmut = c2_nonmut + c3_nonmut) %>%
  dplyr::select(Hugo_Symbol, c1_mut, c1_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust1_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_1_summary <- sig %>%
  left_join(clust1_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  mutate(c1_rate = c1_mut / (c1_mut + c1_nonmut)) %>%
  print()

clust1_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c2_mut + c3_mut) %>%
  mutate(pool_nonmut = c2_nonmut + c3_nonmut) %>%
  dplyr::select(Hugo_Symbol, c1_mut, c1_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust1_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_1_summary <- sig %>%
  left_join(clust1_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  mutate(c1_rate = c1_mut / (c1_mut + c1_nonmut)) %>%
  print()
```

```{r}
clust2_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c1_mut + c3_mut) %>%
  mutate(pool_nonmut = c1_nonmut + c3_nonmut) %>%
  dplyr::select(Hugo_Symbol, c2_mut, c2_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust2_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_2_summary <- sig %>%
  left_join(clust2_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  print()
```

```{r}
clust3_vs_pool <- mut_counts_filtered %>%
  mutate(pool_mut = c1_mut + c2_mut) %>%
  mutate(pool_nonmut = c1_nonmut + c2_nonmut) %>%
  dplyr::select(Hugo_Symbol, c3_mut, c3_nonmut, pool_mut, pool_nonmut) %>%
  print()

sig <- clust3_vs_pool %>%
  nest(-Hugo_Symbol) %>%
  mutate(matrix = map(data, ~matrix(unlist(.x), nrow = 2))) %>% 
  mutate(fisher = map(matrix, ~fisher.test(.x))) %>%
  mutate(stats = map(fisher, ~broom::glance(.x))) %>%
  unnest(stats) %>%
  dplyr::select(Hugo_Symbol, p.value, odds = estimate) %>%
  arrange(p.value)

clust_3_summary <- sig %>%
  left_join(clust3_vs_pool) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::select(Hugo_Symbol, p_adj, p = p.value, everything()) %>%
  print()
```
